---
title: 'Survival On the Titanic'
author: 'Lisa L Stuart'
date: 'February 10, 2017'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Intro
This is my very first Kaggle script!  Looks like this is where a lot of people start to get the hang of it, so I figured I would too.

## Load Libraries
```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('mice') # imputation
library('mi') # imputation
library('randomForest') # classification algorithm
library('gmodels') # for cross tabulation tables
```

## Load Data
```{r, message=FALSE, warning=FALSE}
train <- read.csv('./input/train.csv', stringsAsFactors = F)
test  <- read.csv('./input/test.csv', stringsAsFactors = F)

str(train)
str(test)
full  <- bind_rows(train, test) # bind training & test data

# check data
str(full)
unique(full$Survived) # values are 1, 0, and NA (for the missing Survived row in test)
```

We can see the full dataset contains 1309 observations.  

Variable Name | Description
--------------|-------------
PassengerID   | The ID number of the passenger
Survived      | Survived (1) or died (0) [or NA for the missing values in test]
Pclass        | Passenger's class
Name          | Passenger's name
Sex           | Passenger's sex
Age           | Passenger's age
SibSp         | Number of siblings/spouses aboard
Parch         | Number of parents/children aboard
Ticket        | Ticket number
Fare          | Fare
Cabin         | Cabin
Embarked      | Port of embarkation

# Where are the nulls?
```{r, message=FALSE, warning=FALSE}
sum(is.na(full$PassengerId)) 
sum(is.na(full$Survived)) # corresponds with the number in test set
sum(is.na(full$Pclass)) 
sum(is.na(full$Name)) 
sum(is.na(full$Sex)) 
sum(is.na(full$Age)) #263 - need some fancy imputing here!
sum(is.na(full$SibSp)) 
sum(is.na(full$Parch)) 
sum(is.na(full$Ticket)) 
sum(is.na(full$Fare)) # this shouldn't be too difficult to fill
sum(full$Cabin == "") # yikes, most missing here!
sum(full$Embarked == "") 
```

I liked Megan Risdal's idea of grabbing the title from the name, so let's do that too:

```{r, message=FALSE, warning=FALSE}
# Grab title from passenger names
full$Title <- gsub('(.*, )|(\\..*)', '', full$Name)

# Show title counts by sex
table(full$Sex, full$Title)

# Found these different titles I've never heard of intriguing so did a little Googling.  Don/Dona are used in Spain, Portugal and Italy as a title of honor (such as Sir when knighted in British culture) while Jonkeer is Dutch translating to Young Lord.  Master is for a boy or young man not yet married and Miss is the same for a girl or young woman.
# Hmm, who knew?

# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle (mademoiselle), ms, and mme (madame) accordingly
full$Title[full$Title == 'Mlle']        <- 'Miss' 
full$Title[full$Title == 'Ms']          <- 'Miss'
full$Title[full$Title == 'Mme']         <- 'Mrs' 
full$Title[full$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(full$Sex, full$Title)

# Finally, grab surname from passenger name
full$Surname <- sapply(full$Name,  
                      function(x) strsplit(x, split = '[,.]')[[1]][1])

```

## Family size

```{r}
# Create a family size variable including the passenger themselves
full$FamSize <- full$SibSp + full$Parch + 1

# Create a family variable 
full$Family <- paste(full$Surname, full$FamSize, sep='_')
```

## Plot of Family Size vs Survival Status

```{r, message=FALSE, warning=FALSE}
# Use ggplot2 to visualize the relationship between family size & survival
ggplot(full[1:891,], aes(x = FamSize, fill = factor(Survived))) +
  geom_bar(stat='count', position='dodge') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Family Size') +
  theme_few()
```

Think I'll keep FamilySize numeric for now.

```{r, message=FALSE, warning=FALSE}
numSurvived = sum(train$Survived) # 342
numDied = nrow(train) - numSurvived #549
PropSurvived = numSurvived/nrow(train) # 0.3838384
PropDied = numDied/nrow(train) # 0.6161616
byClass <- group_by(train, Survived, Pclass)
summarise(byClass, train = n())

    
```
Based on these numbers, seems like class made a difference to whether or not you survived.  Let's take a look at a visual to make this more 
clear:

```{r, message=FALSE, warning=FALSE}
# Use ggplot2 to visualize the relationship between class & survival
ggplot(full[1:891,], aes(x = Pclass, fill = factor(Survived))) +
  geom_bar(stat='count', position='dodge') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Class') +
  theme_few()
```
Indeed, class does seem important, but I guess we'll know for sure exactly how important when we get to feature selection a bit later.

## A few tables:
For now let's look at a few tables like survival by sex and embarked by survived to get a feel for a few other relationships.

```{r, message=FALSE, warning=FALSE}
# Table for survival by sex
table(train$Survived, train$Sex)
table(train$Embarked, train$Survived)
```

## Add indicator columns and engineer a few features:
Then let's add a few indicator columns missing age and cabin number, and a few columns for the number of characters in the name and ticket, 
and then the first characters in the cabin and ticket.

```{r, message=FALSE, warning=FALSE}
full$Missing_Age = ifelse(is.na(full$Age), 1, 0)
full$Missing_Cabin = ifelse(full$Cabin == "", 1, 0)
full$NumCharName = nchar(full$Name)
full$NumCharTicket <- nchar(full$Ticket)
full$FirstCharTicket<-factor(sapply(full$Ticket, function(x) strsplit(x, NULL)[[1]][1]))
full$FirstCharCabin<-factor(sapply(full$Cabin, function(x) strsplit(x, NULL)[[1]][1]))
```

# Let's fix what missing

## Missing Embarked

```{r, message=FALSE, warning=FALSE}
# missing from Embarked
which(full$Embarked == "") # passengers 62 and 830
```

```{r results='asis'}
cat(paste('We will infer their values for **embarkment** based on present data that we can imagine may be relevant: 
**passenger class** and **fare**. We see that they paid<b> $', full[c(62, 830), 'Fare'][[1]][1], '</b>and<b> $', full[c(62, 830), 
'Fare'][[1]][2], '</b>respectively and their classes are<b>', full[c(62, 830), 'Pclass'][[1]][1], '</b>and<b>', full[c(62, 830), 
'Pclass'][[1]][2], '</b>. So from where did they embark?'))
```

```{r, message=FALSE, warning=FALSE}
# Get rid of our missing passenger IDs
embark_fare <- full %>%
  filter(PassengerId != 62 & PassengerId != 830)

# Use ggplot2 to visualize embarkment, passenger class, & median fare
ggplot(embark_fare, aes(x = Embarked, y = Fare, fill = factor(Pclass))) +
  geom_boxplot() +
  geom_hline(aes(yintercept=80), 
    colour='red', linetype='dashed', lwd=2) +
  scale_y_continuous(labels=dollar_format()) +
  theme_few()
```

Since both passengers with missing values are first class and the fare they both paid is $80 (which is the average fare for those 
embarking from 'Charbourg' it makes sense to assign them both 'C' for Embarkment.

```{r}
# Since their fare was $80 for 1st class, they most likely embarked from 'C'
full$Embarked[c(62, 830)] <- 'C'
```

## Missing Fare

```{r, message=FALSE, warning=FALSE}
# missing from Fare
which(full$Fare == "") # passenger 1044

# showing the row reveals that this is a 3rd class passenger departing from 'Southampton.'
full[1044, ]
```

So let's look at the distribution of fares for other 3rd class passengers that also departed from Southampton 
(n = `r nrow(full[full$Pclass == '3' & full$Embarked == 'S', ]) - 1`).

```{r, message=FALSE, warning=FALSE}
ggplot(full[full$Pclass == '3' & full$Embarked == 'S', ], 
  aes(x = Fare)) +
  geom_density(fill = '#99d6ff', alpha=0.4) + 
  geom_vline(aes(xintercept=median(Fare, na.rm=T)),
    colour='red', linetype='dashed', lwd=1) +
  scale_x_continuous(labels=dollar_format()) +
  theme_few()
```

Seems fair to replace the NA for this passenger's fare with the median of the other 3rd class passengers who departed from Southampton 
which is $`r  median(full[full$Pclass == '3' & full$Embarked == 'S', ]$Fare, na.rm = TRUE)`.

```{r}
full$Fare[1044] <- median(full[full$Pclass == '3' & full$Embarked == 'S', ]$Fare, na.rm = TRUE)
```

## Missing Age

We know from earlier that there are 263 missing age values

Just for fun, let's try a few different imputation methods and see what their differences are:

### Mice imputation

```{r, message=FALSE, warning=FALSE}
# Make variables factors into factors
factor_vars <- c('PassengerId','Pclass','Sex','Embarked',
                 'Title','Surname','Family', 'FirstCharCabin', 'FirstCharTicket')

full[factor_vars] <- lapply(full[factor_vars], function(x) as.factor(x))

# Set a random seed
set.seed(129)

# Perform mice imputation, excluding certain less-than-useful variables:
mice_mod <- mice(full[, !names(full) %in% c('PassengerId','Name','Ticket','Cabin','Family','Surname','Survived', 'FirstCharCabin', 'FirstCharTicket')], method='rf') 

# Save the complete output 
mice_output <- complete(mice_mod)
```


### mi imputation

```{r, message=FALSE, warning=FALSE}
# impute missing values, using all parameters as default values
mi_imp <- mi(full[ , !names(full) %in% c('PassengerId','Name','Ticket','Cabin','Family','Surname', 'Survived', 'FirstCharTicket', 'FirstCharCabin')], seed = 1234)
```

Now let's compare the results we get with the original distribution of passenger ages to see which looks imputation method looks best.

```{r}
# Plot age distributions
par(mfrow=c(1,2))
hist(full$Age, freq=F, main='Age: Original Data', 
  col='darkgreen', ylim=c(0,0.04))
hist(mice_output$Age, freq=F, main='Age: MICE Output', 
  col='lightgreen', ylim=c(0,0.04))
hist(mi_imp$Age, freq=F, main='Age: mi Output', 
  col='blue', ylim=c(0,0.04))
```

Things look good, so let's replace our age vector in the original data with the output from the `mice` model.







